---
title: "QuakeExplorer"
author: "Tony Gojanovic"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## QuakeExplorer Package Overview

QuakeExplorer is a R package used for the specific purpose of exploring earth quake data from the  NOAA (National Oceanic and Atmospheric Administration) Significant Earthquake dataset.  Data can be obtained from the NOAA at https://www.ngdc.noaa.gov/nndc/struts/form?t=101650&s=1&d=1.  The dataset is historical covering the time span from 2150 BCE to CE 2018  along with location and earthquake statistics such as magnitude, intensity and damage metrics (e.g. deaths) to name but a multitude of measures available for exploration.

QuakeExplorer provide routines for importing, cleaning (tidying) and ploting elements of the data set.

## R Packages

The following R packages are used:

```{r}
library(QuakeExplorer)
library(ggplot2)
library(dplyr)
library(readr)
```

## Dowloading and Importing the NOAA Significant Earthquake Data Set

The earthquake dataset from NOAA can be accessed from https://www.ngdc.noaa.gov/nndc/struts/form?t=101650&s=1&d=1.  From the NOAA site, the data set can be downloaded as a **tab delimiated** format which can be used for importing into R (or any other package suited to read tab delimited files).  

Once the data set is downloaded to a working directory, it will be named **results** but the extention **.csv** must be added.  Then, the following can be used to import the data to an R data frame:

```{r message=FALSE}
import_quake_data("results.csv")
```

